Appendix

### 2024-10-17 manuscript_rmarkdown/manuscript.tex
Update figures to work with ggplot2 and patchwork

\title{Location and Scale-Invariant Power Transformations for
Transforming Data to Normality}
\author{Alex Zwanenburg, Steffen Löck}
\date{2024-09-19}

@@

The \(\lambda\)-parameter is typically optimised using maximum
likelihood estimation under the assumption that the transformed feature
is normally distributed. As noted by Raymaekers and Rousseeuw, this
approach is sensitive to outliers, and robust versions of Box-Cox and
Yeo-Johnson transformations were devised (Raymaekers and Rousseeuw
2021).

@@

Real-world data may contain outliers, to which maximum likelihood
estimation can be sensitive. Their presence may lead to poor
transformations to normality, as shown in Figure
\ref{fig:decreased-normality}. As indicated by Raymaekers and Rousseeuw
(2021), the general aim of power transformations should be to transform
non-outlier data to normality, i.e.~achieve \emph{central normality}. To
achieve this, they devised an iterative procedure to find a robust
estimate of the transformation parameter \(\lambda\). Briefly, this
process requires identifying outliers in the data and weighting such
instances during the optimisation process. Raymaekers and Rousseeuw
(2021) achieve this through weighted maximum likelihood estimation.
However, because this procedure iteratively estimates and updates
\(\lambda\), it can not be used here to simultaneously estimate
\(\lambda\), \(x_0\) and \(s\) for location- and scale-invariant power
transformations. Nonetheless, as a procedure, weighted MLE can be used
for estimating the transformation, shift and scale parameters.

Here, weighted maximum likelihood estimation is based on equations
\ref{eqn:box-cox-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-invariant-log-likelihood}. Compared to Raymaekers
and Rousseeuw (2021), these log-likelihood functions includes additional
terms to accommodate estimation of \(x_0\) and \(s\). The weighted
location- and scale-invariant Box-Cox log-likelihood function is:

@@

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After Raymaekers and
  Rousseeuw (2021),
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) (Huber 1981). Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}

@@

Outlier values were then drawn to randomly replace a fraction of the
values of \(\mathbf{X}_i\). This was repeated \(m_{\text{out}} = 10\)
times, with outlier fractions regularly spaced in \([0.00, 0.10]\). Thus
up to 10 percent of the values could be replaced by outliers. Outlier
values were set according to John Wilder Tukey (1977), as follows. Let
\(x^{*} \sim U\left(-2, 2\right)\). Then the corresponding outlier value
was:

@@

Robust location- and scale-invariant transformations are a suitable
replacement for their conventional counterparts. They demonstrated
robustness against outliers and prevent inaccurate transformations and
potential numerical issues due to location and scale of the distribution
of a feature. This is particularly relevant for automated data
processing, where such issues may go unnoticed. Compared to non-robust
location- and scale-invariant transformations, real-world examples with
outliers showed higher residual errors of transformed features. Robust
transformations seek to minimise residual errors for the central part of
the distribution, instead of the entire distribution, including
outliers. The empirical central normality test indicated that robust
transformations are better able to achieve central normality in the
presence of outliers. However, in a machine learning experiment of 232
real-world datasets that contained at least one numeric feature, we did
not find a meaningful benefit of for -- nor detriment to -- model
performance for location- and scale-invariant power transformations. One
reason may be that numeric features with large location shifts
(\(|\mu| > 1000.0\)) were uncommon. Of the 4886 numeric features in the
232 datasets, 266 (5\%) features in 34 datasets had large location
shifts, of which 200 appeared in just 2 datasets. For the latter two
datasets, the transformation method does not show significant difference
between groups (Friedman test; \(p > 0.05\)).

@@

We assessed several methods for robust power transformation. Methods
that relied on the z-score of the transformed feature or the residual
error yielded worse results than the non-robust method. This is partly
due to the initial choice of threshold parameters. Using different
initial values, closer to the upper limits
(\(\delta_1 = 10.0; \delta_2 = 10.0\)), led to a reduced loss. However,
these values effectively correspond to a non-robust transformation,
where all instances receive the same weight. Underperformance of these
methods could be explained by their use of transformed features for
setting weights. Consequently, the weights change at each iteration in
the MLE optimisation process. This increases local variance in the
log-likelihood function and create local optima that the optimiser may
not handle well. As a consequence, optimal values for transformation
parameter \(\lambda\) might differ, which increases the presented loss
for optimising threshold parameters. Methods that relied on the
empirical probability did not suffer from this issue, as weights
remained fixed during MLE.

We introduced an empirical test for central normality to assess whether
distributions deviate from normality in a way that might require closer
inspection prior to further processing. The empirical test for central
normality differs from other tests for normality, such as the
Shapiro-Wilk test (Shapiro and Wilk 1965), in that the test statistic is
independent of the number of samples. This makes this test more
practical for assessing central normality for larger sample numbers,
where other test may detect inconsequential deviations from normality.

This work has the following limitation. Firstly, we did observe several
numerical stability issues for optimisation criteria other than MLE
(Appendix B). These appear in regions where transformation parameters
would lead to very large or small numbers when using conventional power
transformations. For MLE stability issues were not observed. Secondly,
the empirical central normality test is based on simulations instead of
statistical theory, and relies on a somewhat arbitrary definition of the
central portion of a distribution. Thus, while the test may asses
whether data is sufficiently normally distributed for practical
purposes, it should not be used as a strict test for normality.

@@

\bibitem[\citeproctext]{ref-Box1964-mz}
Box, G E P, and D R Cox. 1964. {``An Analysis of Transformations.''}
\emph{J. R. Stat. Soc. Series B Stat. Methodol.} 26 (2): 211--52.

@@

\bibitem[\citeproctext]{ref-Griffin2018-bf}
Griffin, Maryclare. 2018. \emph{Gnorm: Generalized Normal/Exponential
Power Distribution}. \url{https://CRAN.R-project.org/package=gnorm}.

@@

\bibitem[\citeproctext]{ref-Raymaekers2021-kq}
Raymaekers, Jakob, and Peter J Rousseeuw. 2021. {``Transforming
Variables to Central Normality.''} \emph{Mach. Learn.}, March.
\url{https://doi.org/10.1007/s10994-021-05960-5}.

@@

\bibitem[\citeproctext]{ref-Subbotin1923-qk}
Subbotin, M Th. 1923. {``On the Law of Frequency of Error.''}
\emph{Математический Сборник} 31 (2): 296--301.

@@

\bibitem[\citeproctext]{ref-Tukey1957-rt}
Tukey, John W. 1957. {``On the Comparative Anatomy of
Transformations.''} \emph{Ann. Math. Stat.} 28 (3): 602--32.

\bibitem[\citeproctext]{ref-Tukey1977-xm}
Tukey, John Wilder. 1977. \emph{Exploratory Data Analysis}.
Addison-Wesley Publishing Company.

>>>>

\title{Location and Scale-Invariant Power Transformations for
Transforming Data to Normality}
\author{Alex Zwanenburg, Steffen Löck}
\date{2024-10-17}

@@

The \(\lambda\)-parameter is typically optimised using maximum
likelihood estimation under the assumption that the transformed feature
is normally distributed. As noted by Raymaekers and Rousseeuw, this
approach is sensitive to outliers, and robust versions of Box-Cox and
Yeo-Johnson transformations were devised (Raymaekers and Rousseeuw
2024).

@@

Real-world data may contain outliers, to which maximum likelihood
estimation can be sensitive. Their presence may lead to poor
transformations to normality, as shown in Figure
\ref{fig:decreased-normality}. As indicated by Raymaekers and Rousseeuw
(2024), the general aim of power transformations should be to transform
non-outlier data to normality, i.e.~achieve \emph{central normality}. To
achieve this, they devised an iterative procedure to find a robust
estimate of the transformation parameter \(\lambda\). Briefly, this
process requires identifying outliers in the data and weighting such
instances during the optimisation process. Raymaekers and Rousseeuw
(2024) achieve this through weighted maximum likelihood estimation.
However, because this procedure iteratively estimates and updates
\(\lambda\), it can not be used here to simultaneously estimate
\(\lambda\), \(x_0\) and \(s\) for location- and scale-invariant power
transformations. Nonetheless, as a procedure, weighted MLE can be used
for estimating the transformation, shift and scale parameters.

Here, weighted maximum likelihood estimation is based on equations
\ref{eqn:box-cox-invariant-log-likelihood} and
\ref{eqn:yeo-johnson-invariant-log-likelihood}. Compared to Raymaekers
and Rousseeuw (2024), these log-likelihood functions includes additional
terms to accommodate estimation of \(x_0\) and \(s\). The weighted
location- and scale-invariant Box-Cox log-likelihood function is:

@@

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After Raymaekers and
  Rousseeuw (2024),
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) (Huber 1981). Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}

@@

Outlier values were then drawn to randomly replace a fraction of the
values of \(\mathbf{X}_i\). This was repeated \(m_{\text{out}} = 10\)
times, with outlier fractions regularly spaced in \([0.00, 0.10]\). Thus
up to 10 percent of the values could be replaced by outliers. Outlier
values were set according to John W. Tukey (1977), as follows. Let
\(x^{*} \sim U\left(-2, 2\right)\). Then the corresponding outlier value
was:

@@

Robust location- and scale-invariant transformations are a suitable
replacement for their conventional counterparts. They demonstrated
robustness against outliers and prevent inaccurate transformations and
potential numerical issues due to location and scale of the distribution
of a feature. This is particularly relevant for automated data
processing, where such issues may go unnoticed. Compared to non-robust
location- and scale-invariant transformations, real-world examples with
outliers showed higher residual errors of transformed features. Robust
transformations seek to minimise residual errors for the central part of
the distribution, instead of the entire distribution, including
outliers. The empirical central normality test indicated that robust
transformations are better able to achieve central normality in the
presence of outliers. However, in a machine learning experiment of 232
real-world datasets that contained at least one numeric feature, we did
not find a meaningful benefit -- nor detriment -- to model performance
for location- and scale-invariant power transformations. One reason may
be that numeric features with large location shifts (\(|\mu| > 1000.0\))
were uncommon. Of the 4886 numeric features in the 232 datasets, 266
(5\%) features in 34 datasets had large location shifts, of which 200
appeared in just 2 datasets. For the latter two datasets, the
transformation method did not show significant difference between groups
(Friedman test; \(p > 0.05\)).

@@

We assessed several methods for robust power transformation. Methods
that relied on the z-score of the transformed feature or the residual
error yielded worse results than the non-robust method. This is partly
due to the initial choice of threshold parameters. Using different
initial values, closer to the upper limits
(\(\delta_1 = 10.0; \delta_2 = 10.0\)), led to a reduced loss. However,
these values effectively correspond to a non-robust transformation,
where all instances receive the same weight. Underperformance of these
methods could be explained by their use of transformed features for
setting weights. Consequently, the weights change at each iteration in
the MLE optimisation process. This increases local variance in the
log-likelihood function and creates local optima that the optimiser may
not handle well. As a consequence, optimal values for transformation
parameter \(\lambda\) might differ, which increases the presented loss
for optimising threshold parameters. Methods that relied on the
empirical probability did not suffer from this issue, as weights
remained fixed during MLE.

We introduced an empirical test for central normality to assess whether
distributions deviate from normality in a way that might require closer
inspection prior to further processing. The empirical test for central
normality differs from other tests for normality, such as the
Shapiro-Wilk test (Shapiro and Wilk 1965), in that the test statistic is
independent of the number of samples. This makes this test more
practical for assessing central normality for larger sample numbers,
where other tests may detect inconsequential deviations from normality.

This work has the following limitations. Firstly, we did observe several
numerical stability issues for optimisation criteria other than MLE
(Appendix B). These appear in regions where transformation parameters
would lead to very large or small numbers when using conventional power
transformations. For MLE stability issues were not observed. Secondly,
the empirical central normality test is based on simulations instead of
statistical theory, and relies on a somewhat arbitrary definition of the
central portion of a distribution. Thus, while the test may asses
whether data is sufficiently normally distributed for practical
purposes, it should not be used as a strict test for normality.

@@

\bibitem[\citeproctext]{ref-Box1964-mz}
Box, G E P, and D R Cox. 1964. {``An Analysis of Transformations.''}
\emph{J. R. Stat. Soc. Series B Stat. Methodol.} 26 (2): 211--52.
\url{https://doi.org/10.1111/J.2517-6161.1964.TB00553.X}.

@@

\bibitem[\citeproctext]{ref-Griffin2018-bf}
Griffin, Maryclare. 2018. {``Gnorm: Generalized Normal/Exponential Power
Distribution.''} The R Foundation.
\url{https://doi.org/10.32614/cran.package.gnorm}.

@@

\bibitem[\citeproctext]{ref-Raymaekers2024-zf}
Raymaekers, Jakob, and Peter J Rousseeuw. 2024. {``Transforming
Variables to Central Normality.''} \emph{Mach. Learn.} 113 (8):
4953--75. \url{https://doi.org/10.1007/s10994-021-05960-5}.

@@

\bibitem[\citeproctext]{ref-Subbotin1923-qk}
Subbotin, M Th. 1923. {``On the Law of Frequency of Error.''} \emph{Mat.
Sb.} 31 (2): 296--301.

@@

\bibitem[\citeproctext]{ref-Tukey1957-rt}
Tukey, John W. 1957. {``On the Comparative Anatomy of
Transformations.''} \emph{Ann. Math. Stat.} 28 (3): 602--32.
\url{https://doi.org/10.1214/AOMS/1177706875}.

\bibitem[\citeproctext]{ref-Tukey1977-xm}
---------. 1977. \emph{Exploratory Data Analysis}. Addison-Wesley
Publishing Company.

### 2024-10-17 manuscript_rmarkdown/manuscript_appendix.Rmd
Update figures to work with ggplot2 and patchwork

---
title: "Appendix"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
papersize: a4
bibliography: refs.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \DeclareMathOperator*{\argmax}{argmax}
  - \DeclareMathOperator*{\argmin}{argmin}
  - \DeclareMathOperator{\sgn}{sgn}
---

@@

```{r shifted-distributions-appendix, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Invariant power transformation produces transformation parameters that are invariant to location and scale. ",
  "Samples were drawn from normal, right-skewed and left-skewed distributions, respectively, which then underwent a shift $d_{\\text{shift}}$ or multiplication by $d_{\\text{scale}}$. ",
  "Estimates of the transformation parameter $\\lambda$ for the conventional power transformations show strong dependency on the overall location and scale of the distribution and the optimisation criterion, ",
  "whereas estimates obtained for the location- and scale-invariant power transformations are constant."
)

>>>>

---
title: "Appendix"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \DeclareMathOperator*{\argmax}{argmax}
  - \DeclareMathOperator*{\argmin}{argmin}
  - \DeclareMathOperator{\sgn}{sgn}
---

@@

```{r shifted-distributions-appendix, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Invariant power transformation produces transformation parameters that are invariant to location and scale. ",
  "Samples were drawn from normal, right-skewed and left-skewed distributions, respectively, which then underwent a shift $d_{\\text{shift}}$ or multiplication by $d_{\\text{scale}}$. ",
  "Estimates of the transformation parameter $\\lambda$ for the conventional power transformations show strong dependency on the overall location and scale of the distribution and the optimisation criterion, ",
  "whereas estimates obtained for the location- and scale-invariant power transformations are constant. ",
  "For location- and scale-invariant power transformations, the Anderson-Darling criterion leads to unstable estimates of $\\lambda$ for skewed distributions, possibly due to large weights being assigned to samples at the upper and lower ends of the distribution." 
)
